\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-11-10 lang-design.tex
edits in §3.1

We present the key ingredients of our design as an extension to the Coq
proof assistant,
though we believe it could be adapted to other proof assistants such as Lean.
We call our design and implementation \Lang (\underline{f}amily
\underline{\smash{po}}lymorphism for a \underline{\smash{p}}roof assistant).
In this section, we focus on the language design of \Lang.
\cref{sec:coqimpl} describes its implementation as a Coq plugin.

@@

In family "STLCFix", inductive type "tm" \emph{further binds} the "tm" type
in family "STLC". It is generated by five constructors: four inherited from
"STLC" and a fifth called "tm_fix".

Crucially, the types of the constructors are polymorphic to their enclosing family:
the types mention "tm", and the meaning of "tm" is \emph{late bound}.
For example, although "tm_app" is defined in the base family,
it can be used to construct applications of fixpoints in the derived family:
%such as \lsti[mathescape]{tm_app (tm_fix "f" (tm_abs "x" $\dadada$)) tm_unit},
its type \lsti{tm -> tm -> tm} says "tm_app" can be applied to any
Gallina terms of type "tm",
which in family "STLCFix" include those constructed from "tm_fix".

@@

As an example, consider the substitution function "subst", defined using
\lsti!FRecursion!.
The \lsti{on} clause specifies that recursion is over "tm".
The \lsti{motive} clause suggests that the recursive function being
defined has type \lsti{tm -> id -> tm -> tm}.
%
The "subst" function in "STLC" is further bound by the "subst" in "STLCFix":
the four cases from "STLC" are automatically inherited and reused,
with "STLCFix" retroactively adding a fifth case to form the new "subst" function.
%
For exhaustivity, it is a static error if the programmer fails to
further bind "subst" and define this fifth case.
The type system does this check by examining if the inductive type "tm",
over which "subst" is recursively defined, is further bound in the same family.

@@

\begin{lstlisting}[basicstyle=\fontsize{8}{9}\ttfamily]
\forall G t T, hasty G t T -> \forall G', includedin G G' -> hasty G' t T.
\end{lstlisting}

@@

\noindent
The programmer can use tactic programming to discharge the goal.
%The idiomatic programming experience with Coq is largely preserved.
Because the lemma is by induction on "hasty" and because "STLCFix" adds
a constructor to "hasty", the programmer is required to handle this extra case
in the proof of "substlem" in family "STLCFix".

>>>>

We present the key ingredients of our design as an extension to the Coq
proof assistant,
though we believe the design could be adapted to other proof assistants such as Lean.
We call our design and implementation \Lang (\underline{f}amily
\underline{\smash{po}}lymorphism for a \underline{\smash{p}}roof assistant).
In this section, we focus on the language design of \Lang.
\cref{sec:coqimpl} describes its implementation as a Coq plugin.

@@

In family "STLCFix", inductive type "tm" \emph{further binds} the "tm" type
in family "STLC". It has five constructors: four inherited from
"STLC" and a fifth called "tm_fix".

Crucially, the meanings of \lsti{tm} and its constructors are
\emph{late bound}, depending on the family in which they are referenced.
Consider \lsti{tm_app}.
It is defined in family \lsti{STLC} and is thus unaware of \lsti{tm_fix}.
Yet in family \lsti{STLCFix}, we can use \lsti{tm_app} to construct
applications of fixpoints, as in
\lsti[mathescape]{tm_app (tm_fix "f" $t_1$) $t_2$}.
This use is justified by \lsti{tm_app}'s type,
\lsti[mathescape]{tm$\,$->$\,$tm$\,$->$\,$tm}.
It allows \lsti{tm_app} to be applied to anything of type \lsti{tm},
which in family "STLCFix" include those constructed from "tm_fix".

@@

As an example, consider the substitution function "subst", defined using
\lsti!FRecursion!.
The \lsti{on} clause specifies that recursion is over "tm".
The \lsti{motive} clause suggests that the recursive function being
defined has type \lsti{tm -> id -> tm -> tm}.
%
The "subst" function in "STLC" is further bound by the "subst" in "STLCFix":
the four cases from "STLC" are automatically inherited and reused,
with "STLCFix" retroactively adding a fifth case to form the new "subst" function.
%
For exhaustivity, it is a static error if the programmer fails to
further bind "subst" and define this fifth case.
The type system does this check in family \lsti{STLCFix}
by examining if the inductive type "tm",
over which "subst" is recursively defined, is further bound in the same family.

@@

\begin{centered}
\begin{minipage}{.63\textwidth}
\begin{lstlisting}[basicstyle=\fontsize{8}{9}\ttfamily]
\forall G t T, hasty G t T -> \forall G', includedin G G' -> hasty G' t T.
\end{lstlisting}
\end{minipage}
\end{centered}

@@

\noindent
The programmer can use tactic programming to discharge the goal.
%The idiomatic programming experience with Coq is largely preserved.
Because the lemma is by induction on "hasty" and because "STLCFix" adds
a constructor to "hasty", the programmer is required in family "STLCFix" to
extend the proof of "substlem" to handle this extra case.
